Conservation and losses of avian non-coding RNA loci

### 2014-05-23 paper/manuscript.tex
Updated supplementary images.
Updated intro.

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive. Consequently, for this work we have concentrated on
bioinformatic approaches. Primarily, homology based methods, namely
covariance models (CMs). These remain state of the art for ncRNA
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

There is substantial gain and loss of lncRNAs
\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
many of these observations are due to limited bioinformatic sequence
alignment tools (these generally fail align correctly below 60-50\%
sequence identity \cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or
due to genuine gains and losses. Nevertheless, sequence conservation,
generally speaking, provides useful evidence for gene and function
conservation.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:1}).

>>>>

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive in terms of researcher-time and
finances. Consequently, for this work we have concentrated on
bioinformatic approaches. However, we do validate the majority of
these predictions using RNA-seq. Primarily, homology based methods,
namely covariance models (CMs). These remain state of the art for
ncRNA bioinformatic analyses
\cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity can be limited for
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

@@

There is substantial gain and loss of lncRNAs and other ncRNA loci
over evolutionary time
\cite{Cabili:2011,Kutter:2012,Hoeppner:2012}. It is difficult to
assess how many of these ``gains'' and ``losses'' are due to limited
bioinformatic sequence alignment tools (these generally fail align
correctly below 60-50\% sequence identity
\cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or due to genuine
gains and losses. Nevertheless, sequence conservation, generally
speaking, provides useful evidence for gene and function conservation.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1,
SOX2OT and ST7-OT3 lncRNAs have multiple, well conserved RNA-domains
(See Figure~\ref{fig:1}). The syntenic ordering of these seven lncRNAs
and the flanking genes are also preserved between the human and
chicken genomes (data not shown). We illustrate this in detail for the
HOTAIRM1 lncRNA (See Supplemental Figures~13\&14).
